
---
title: |
       | Introduction to Bayesian pharmacometric data analysis using NONMEM
       |
       | Hands-on 1: MAP Bayes example
author: "Metrum Research Group"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    css: styles/styles.css
    highlight: pygments
    theme: flatly
---

# NONMEM workflow example

## PK-PD modeling of time-averaged biomarker and PK data
 
Phase 1 single dose study in healthy volunteers

* Parallel dose-escalation design
* 8 subjects per dose arm
* Single doses of ME-2
    + Placebo, 1.25, 5, 10, 15, 20, 30, 40, 60 and 80 mg
* PK: plasma concentrations of parent drug
* Biomarker: ex vivo inhibition of factor Xa activity in plasma
    + PK and biomarker measured at 0, 0.083, 0.167, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 18 and 24 hours after dose.
    
Hands-on exercise:

* Model relationship between time-averaged factor Xa inhibition and time-averaged ME-2 plasma concentrations

## Model
  
Sigmoid Emax model relating time-averaged \% inhibition of factor Xa activity to time-averaged ME-2 plasma concentration in the $i^{th}$ subject:
\begin{eqnarray*}
\overline{E}_{24,i} &\sim& N\left(\hat{E}_{24,i},\sigma\right) \\
\hat{E}_{24,i} &=& \frac{E_{max} \overline{c}_{24,i}^\gamma}{EC_{50}^\gamma + \overline{c}_{24,i}^\gamma}
\end{eqnarray*}

Weakly informative prior distributions:
  \begin{align*}
    \text{logit}\left(\frac{E_{max}}{100}\right) &\sim N\left(0,
                                                   \sqrt{2}\right) \\
    \log\left(EC_{50}\right)  &\sim N\left(\log(245), 1\right) \\
\log\left(\gamma\right) &\sim N\left(\log(2.1), \sqrt{0.5}\right) \\
\log\left(\sigma\right) &\sim N\left(\log(10), 1\right)
  \end{align*}

---

## Setup R environment

```{r setup, results = 'hide'}

rm(list = ls())
gc()

modelName <- "fxaInhibitAvgMAP"
scriptName <- paste(modelName, "Rmd", sep = ".")

## Relative paths assuming the working directory is the script directory
## containing this script
scriptDir <- getwd()
projectDir <- dirname(scriptDir)
dataDir <- file.path(projectDir, "data")
modelDir <- file.path(projectDir, "model")
outDir <- file.path(modelDir, modelName)
figDir <- file.path(projectDir, "deliv", "figure", modelName)
tabDir <- file.path(projectDir, "deliv", "table", modelName)
invisible(dir.create(figDir, recursive = TRUE))
invisible(dir.create(tabDir, recursive = TRUE))


library(metrumrg)
suppressMessages(library(tidyverse))
library(gridExtra)
library(parallel)
library(knitr)
library(kableExtra)
library(qapply)
library(PKPDmisc)

knitr::opts_chunk$set(
  comment = '.', 
  fig.height = 5, 
  fig.width = 9,
  eval.after = 'fig.cap',
  message = FALSE,
  fig.path = file.path(figDir, paste(modelName, "_", sep = ""))
)



## Go back to default ggplot2 theme that was overridden by bayesplot
theme_set(theme_gray())

source(file.path(scriptDir, "tools", "stanTools2.R"))
source(file.path(scriptDir, "tools", "functions.R"))

options(mc.cores = parallel::detectCores())

set.seed(11191951) 
set.seed(10271998) 

```

## The NONMEM model stub

```{r nonmemModel}

modelFile <- file.path(modelDir, paste(modelName, "stub.ctl", sep = ""))
modelText <- readLines(modelFile)
cat(modelText, sep = "\n")

```

## Assemble initial estimates, priors and table specs

```{r initsPriors}

## create initial estimates
geninit <- function(){
    paste(c(
    "; Initial values of THETA",
    "$THETA",
    paste(rnorm(1, logit(90/100), 0.25), "     ; EMAX = 100 * EXPIT(THETA(1))"),
    paste(rnorm(1, log(120), 0.25), " ; EC50 = EXP(THETA(2))"),
    paste(rnorm(1, log(1), 0.25), "   ; GAMMA = EXP(THETA(3))"),
    paste(rnorm(1, log(5.9), 0.25), "  ; SD = (THETA(4)) "),
    "$OMEGA DIAG(4);INITIAL values of OMEGAs",
    "(0 FIX)",
    "(0 FIX)",
    "(0 FIX)",
    "(0 FIX)",
    "$SIGMA ;INITIAL values of SIGMAs",
    "(1 FIX)",
    sep = "\n"))
}


## Set parameters of the prior distribution
priors <- paste(c("$PRIOR NWPRI",
                  "$THETAP          ; Prior information of THETAS",
                  "(0 FIX)      ;  THETA(1) EMAX",
                  "(5.5 FIX)      ;  THETA(2) EC50",
                  "(0.75 FIX)      ;  THETA(3) gamma",
                  "(2.3 FIX)      ; THETA(4) SD",
                  "$THETAPV BLOCK(4)     ;  variances for priors on THETAS (var-cov)",
                  "2.0 FIX ; EMAX almost uniform",
                  "0.00 1  ; EC50",
                  "0.00 0.00 0.50  ; gamma",
                  "0.00 0.00 0.00 1   ; SD"
),
sep = "\n")

## Table statements
tables <- paste(c("$TABLE ID EVID CONC DV PRED",
                  "NOPRINT ONEHEADER FILE=./.tab",
                  "$TABLE ID EMAX EC50 GAMMA SD",
                  "NOPRINT ONEHEADER FILE=./par.tab"),
                sep = "\n")

```

## Run NONMEM

``` {r runStan, results = 'hide'}

if(!file.exists(file.path(modelDir, modelName))) 
  dir.create(file.path(modelDir, modelName))
runOpt(modelName, modelDir, priors, tables, method = 0, mode = "nonpara", grid = TRUE)

```

```{r NONMEMResults}

nmLstFile <- file.path(modelDir, modelName, modelName, paste(modelName,  ".lst", sep = ""))
nmLstText <- readLines(nmLstFile)
cat(nmLstText, sep = "\n")


```

```{r EstimateTable}

ptab = PKPDmisc::read_nonmem(path = file.path(modelDir, modelName, modelName, 
                                             paste(modelName, "par.tab", sep = "")))

ptab %>%
  slice(1) %>%
  select(EMAX:SD) %>%
  kable(caption = "Summary of model parameter estimates.") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```



```{r MAPplots}

tab = PKPDmisc::read_nonmem(path = file.path(modelDir, modelName, modelName, 
                                             paste(modelName, ".tab", sep = "")))

p1 <- ggplot(tab, aes(x = CONC, y = DV))
p1 + geom_point() +
    geom_line(aes(x = CONC, y = PRED)) +
    labs(x = "average ME-2 plasma concentration (ng/mL)",
         y = "average factor Xa inhibition (%)") +
    theme(text = element_text(size = 12), axis.text = element_text(size = 12),
              legend.position = "none", strip.text = element_text(size = 8))   


```



```{r theEnd}

sessionInfo()

```
