
---
title: |
       | Introduction to Bayesian pharmacometric data analysis using NONMEM
       |
       | Hands-on 1: METHOD = NUTS example
author: "Metrum Research Group"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    css: styles/styles.css
    highlight: pygments
    theme: flatly
---

# NONMEM workflow example

## PK-PD modeling of time-averaged biomarker and PK data
 
Phase 1 single dose study in healthy volunteers

* Parallel dose-escalation design
* 8 subjects per dose arm
* Single doses of ME-2
    + Placebo, 1.25, 5, 10, 15, 20, 30, 40, 60 and 80 mg
* PK: plasma concentrations of parent drug
* Biomarker: ex vivo inhibition of factor Xa activity in plasma
    + PK and biomarker measured at 0, 0.083, 0.167, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 18 and 24 hours after dose.
    
Hands-on exercise:

* Model relationship between time-averaged factor Xa inhibition and time-averaged ME-2 plasma concentrations

## Model
  
Sigmoid Emax model relating time-averaged \% inhibition of factor Xa activity to time-averaged ME-2 plasma concentration in the $i^{th}$ subject:
\begin{eqnarray*}
\overline{E}_{24,i} &\sim& N\left(\hat{E}_{24,i},\sigma\right) \\
\hat{E}_{24,i} &=& \frac{E_{max} \overline{c}_{24,i}^\gamma}{EC_{50}^\gamma + \overline{c}_{24,i}^\gamma}
\end{eqnarray*}

Weakly informative prior distributions:
  \begin{align*}
    \text{logit}\left(\frac{E_{max}}{100}\right) &\sim N\left(0,
                                                   \sqrt{2}\right) \\
    \log\left(EC_{50}\right)  &\sim N\left(\log(245), 1\right) \\
\log\left(\gamma\right) &\sim N\left(\log(2.1), \sqrt{0.5}\right) \\
\log\left(\sigma\right) &\sim N\left(\log(10), 1\right)
  \end{align*}
---

## Setup R environment

```{r setup, results = 'hide'}

rm(list = ls())
gc()

modelName <- "fxaInhibitAvgNUTS"
scriptName <- paste(modelName, "Rmd", sep = ".")
fitModel <- FALSE

## Relative paths assuming the working directory is the script directory
## containing this script
scriptDir <- getwd()
projectDir <- dirname(scriptDir)
dataDir <- file.path(projectDir, "data")
modelDir <- file.path(projectDir, "model")
outDir <- file.path(modelDir, modelName)
figDir <- file.path(projectDir, "deliv", "figure", modelName)
tabDir <- file.path(projectDir, "deliv", "table", modelName)
invisible(dir.create(figDir, recursive = TRUE))
invisible(dir.create(tabDir, recursive = TRUE))


library(metrumrg)
suppressMessages(library(rstan))
suppressMessages(library(bayesplot))
suppressMessages(library(tidyverse))
library(gridExtra)
library(parallel)
library(knitr)
library(kableExtra)
library(qapply)
library(PKPDmisc)

knitr::opts_chunk$set(
  comment = '.', 
  fig.height = 5, 
  fig.width = 9,
  eval.after = 'fig.cap',
  message = FALSE,
  fig.path = file.path(figDir, paste(modelName, "_", sep = ""))
)


## Go back to default ggplot2 theme that was overridden by bayesplot
theme_set(theme_gray())

source(file.path(scriptDir, "tools", "stanTools2.R"))
source(file.path(scriptDir, "tools", "functions.R"))

options(mc.cores = parallel::detectCores())

set.seed(11191951) 
set.seed(10271998) 

```

## The NONMEM model stub

```{r nonmemModel}

modelFile <- file.path(modelDir, paste(modelName, "stub.ctl", sep = ""))
modelText <- readLines(modelFile)
cat(modelText, sep = "\n")

```

## Assemble initial estimates, priors and table specs

```{r initsPriors}

## create initial estimates
geninit <- function(){
    paste(c(
    "; Initial values of THETA",
    "$THETA",
    paste(rnorm(1, logit(90/100), 0.25), "     ; EMAX = 100 * EXPIT(THETA(1))"),
    paste(rnorm(1, log(120), 0.25), " ; EC50 = EXP(THETA(2))"),
    paste(rnorm(1, log(1), 0.25), "   ; GAMMA = EXP(THETA(3))"),
    paste(rnorm(1, log(5.9), 0.25), "  ; SD = (THETA(4)) "),
    "$OMEGA DIAG(4);INITIAL values of OMEGAs",
    "(0 FIX)",
    "(0 FIX)",
    "(0 FIX)",
    "(0 FIX)",
    "$SIGMA ;INITIAL values of SIGMAs",
    "(1 FIX)",
    sep = "\n"))
}

## Set parameters of the prior distribution
priors <- paste(c("$PRIOR NWPRI",
                  "$THETAP          ; Prior information of THETAS",
                  "(0 FIX)      ;  THETA(1) EMAX",
                  "(5.5 FIX)      ;  THETA(2) EC50",
                  "(0.75 FIX)      ;  THETA(3) gamma",
                  "(2.3 FIX)      ; THETA(4) SD",
                  "$THETAPV BLOCK(4)     ;  variances for priors on THETAS (var-cov)",
                  "2.0 FIX ; EMAX almost uniform",
                  "0.00 1  ; EC50",
                  "0.00 0.00 0.50  ; gamma",
                  "0.00 0.00 0.00 1   ; SD"
),
sep = "\n")

## Table statements
tables <- paste(c("$TABLE ID EVID CONC DV PRED",
                  "NOPRINT ONEHEADER FILE=./.tab",
                  "$TABLE ID EMAX EC50 GAMMA SD",
                  "NOPRINT ONEHEADER FILE=./par.tab"),
                sep = "\n")

```

## Run NONMEM

``` {r runNONMEM, results = 'hide'}

nChains <- 4
nPost <- 500 ## Number of post-burn-in samples per chain after thinning
nBurn <- 500 ## Number of burn-in samples per chain after thinning
nThin <- 1

seed = sample(10000, size = nChains)
if(fitModel){
  if(!file.exists(file.path(modelDir, modelName))) 
    dir.create(file.path(modelDir, modelName))
  runChains <- mclapply(1:nChains, runChain,
                      modelName = modelName, modelDir = modelDir, 
                      priors = priors, tables = tables,
                      nPost = nPost, nBurn = nBurn, nThin = nThin, 
                      seed = seed, print = 100,
                      OLKJDF = 0, OVARF = 1, AUTO = 2,
                      grid = TRUE, method = "NUTS", mode = "nonpara")
  }

```


## read in data file
```{r getData}

dataFile <- "fxaInhAvg.csv"
nmData <- read.csv(file.path(dataDir, dataFile), as.is = TRUE)

```

## Get population parameters

```{r getPopParam}

chains <- 1:nChains
##chains <- c(1, 3, 4)

nChains2 <- length(chains)

## Read in posterior distributions by chain and add a column called chain
popParameters = map(chains, function(thisChain){
  param <- data.table::fread(file.path(modelDir, modelName, paste(modelName, ".", thisChain, sep = ""), 
                                       "par.txt"), data.table = FALSE) 
  names(param) <- c("iteration",
                    "THETA1", "THETA2", "THETA3", "THETA4")
  param %>% mutate(chain = thisChain)
}) %>% bind_rows %>%
  mutate(sample = 1:n())

## Calculate more interpretable paraneters
popParameters2 <- popParameters %>%
  mutate(EMAX = 100*plogis(THETA1),
         EC50 = exp(THETA2),
         GAMMA = exp(THETA3),
         sigma = exp(THETA4)) %>%
  select(EMAX,EC50,GAMMA,sigma,chain, iteration, sample)

## Convert to 3-D array with dims = {iterations, chains, parameters}. 
popParArray <- array(double(nPost * nChains2 * (ncol(popParameters2) - 3)), 
                     dim = c(nPost, nChains2, ncol(popParameters2) - 3), 
                     dimnames =  list(NULL, NULL, setdiff(names(popParameters2), c("chain", "iteration", "sample"))))
for(iChain in 1:nChains2){
  popParArray[,iChain,] <- popParameters2 %>%
    filter(chain == chains[iChain],
           iteration > 0) %>%
    select(-chain, -iteration, -sample) %>%
    as.matrix
}

```


## MCMC diagnostics and posterior distributions of parameters

``` {r parameters, fig.cap = caption}

options(bayesplot.base_size = 12,
        bayesplot.base_family = "sans")
color_scheme_set(scheme = "brightblue")
myTheme <- theme(text = element_text(size = 12), 
                 axis.text = element_text(size = 12))

plot_mcmcHistory <- mcmcHistory(popParArray,
                                 nParPerPage = 5, myTheme = myTheme)
plot_mcmcDensityByChain <- mcmcDensity(popParArray, 
                                        nParPerPage = 16, byChain = TRUE, 
                                        myTheme = theme(text = element_text(size = 12), 
                                                        axis.text = element_text(size = 10)))
plot_mcmcDensity <- mcmcDensity(popParArray, nParPerPage = 16, 
                                 myTheme = theme(text = element_text(size = 12), 
                                                 axis.text = element_text(size = 10)))

plot_mcmcHistory
plot_mcmcDensityByChain
plot_mcmcDensity
mcmc_pairs(popParArray[,,!grepl("^OMEGA", dimnames(popParArray)[[3]])])

caption <- paste("PK-PD modeling of time-averaged biomarker and PK data example:",
      c(rep("MCMC trace plots.", length(plot_mcmcHistory)),
        rep("Univariate marginal densities by chain.",
            length(plot_mcmcDensityByChain)),
        rep("Univariate marginal densities.",
            length(plot_mcmcDensityByChain)),
        rep("Bivariate marginal densities.", 2)))

ptable <- monitor(popParArray, 
                  warmup = 0, print = FALSE) %>% 
  as.matrix() %>%
  formatC(3) %>%
  as.data.frame
write.csv(ptable, file = file.path(tabDir, paste(modelName, 
                                                 "ParameterTable.csv", sep = "")))

ptable %>% 
  rename(SEmean = se_mean, SD = sd, pct2.5 = "2.5%", pct25 = "25%", median = "50%",
         pct75 = "75%", pct97.5 = "97.5%", Neff = "n_eff") %>%
  mutate(parameter = rownames(.), "95% CI" = paste("(", pct2.5, ", ", pct97.5, ")", 
                                                   sep = "")) %>%
  select(parameter, mean, SD, median, "95% CI", Neff, Rhat) %>%
  kable(caption = "Summary of model parameter estimates.") %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

```


```{r autocorrelation, fig.cap=caption}

autcorr_plot = popParameters2 %>% tidyr::gather(key=parameter, value=value,-c(chain,iteration,sample)) %>%
  select(-c(sample,iteration)) %>%
  group_by(chain,parameter) %>%
  nest() %>%
  mutate(acf = pmap(list(chain,parameter,data), .f = function(.chain,.par,.df) {
    x = as.numeric(.df$value)
    y = acf(x, plot=FALSE, lag.max=20)
    data.frame(lag=as.numeric(y$lag), acf=as.numeric(y$acf))
  })) %>%
  select(-data) %>%
  unnest() %>%
  ggplot(aes(x=lag,y=acf,fill=chain)) +
  geom_bar(stat='identity') +
  facet_grid( parameter ~ chain) +
  labs(x='Lag', y='Autocorrelation')

caption <- paste("PK-PD modeling of time-averaged biomarker and PK data example:",
      "Autocorrelation plots.")

autcorr_plot
```

## Construct design data frame

```{r design}

## For now just use the NONMEM data set

design <- nmData %>%
  filter(C == ".") %>%
  select(-c(EVID,CMT,C))


```

## mrgsolve model code for posterior predictions

```{r model code}

library(mrgsolve)

indCode =
  '
$PLUGIN mrgx

$PARAM 
THETA1 = 5.65
THETA2 = 5.60
THETA3 = 4.29
THETA4 = -1.33
CONC = 10

$PRED

double EMAX = 100 / (1 + exp(-THETA1));
double EC50 = exp(THETA2);
double GAMMA = exp(THETA3);
double sigma = exp(THETA4);

double eps = mrgx::rnorm(0.0, sigma,-100, 100);

double RESP = EMAX * pow(CONC,GAMMA) / (pow(EC50,GAMMA) + pow(CONC,GAMMA));
double simDV = RESP + eps;

$CAPTURE RESP simDV
'

## Compile model
fxaMod <- mcode("FXAmodel", indCode)

if(FALSE){
  ## Test runs
  out <- fxaMod %>%
    data_set(design) %>%
    carry_out(DOSE,CONC) %>%
    mrgsim
  }

```


## Population predictions

```{r popSim}

samples <- unique(popParameters$sample)
popSim1 <- mclapply(samples,
                    function(thisSample){
                      fxaMod %>%
                        data_set(design) %>%
                        param(popParameters %>% 
                                filter(sample == thisSample)) %>%
                        carry_out(sample, ID, DV, CONC, RESP, simDV, DOSE,chain,sample) %>%
                        mrgsim %>%
                        as.tbl
                    }) %>%
  bind_rows

```

## Posterior predictive distributions

``` {r PPC, results = 'asis'}

## Combined plots

popPred <- popSim1 %>%
  group_by(CONC) %>%
  summarize(lbPop = quantile(RESP, probs = 0.05, na.rm = TRUE),
            medianPop = quantile(RESP, probs = 0.5, na.rm = TRUE),
            ubPop = quantile(RESP, probs = 0.95, na.rm = TRUE),
            lbInd = quantile(simDV, probs = 0.05, na.rm = TRUE),
            medianInd = quantile(simDV, probs = 0.5, na.rm = TRUE),
            ubInd = quantile(simDV, probs = 0.95, na.rm = TRUE))

predAll <- popPred %>%
  left_join(design) %>%
  mutate(DV = suppressWarnings(as.numeric(DV)))

#forms <- sort(unique(predAll$FORM))
plot_PPC <- predAll %>% ggplot(aes(x = CONC, y = DV)) + 
  geom_line(aes(x = CONC, y = medianInd, 
                color = "prediction interval")) +
  geom_ribbon(aes(ymin = lbInd, ymax = ubInd, 
                  fill = "prediction interval"), 
              alpha = 0.25) +
  geom_line(aes(x = CONC, y = medianPop, 
                color = "credible interval")) +
  geom_ribbon(aes(ymin = lbPop, ymax = ubPop, 
                  fill = "credible interval"), 
              alpha = 0.25) +
  scale_color_brewer(name  ="prediction",
                     breaks=c("prediction interval", "credible interval"),
                     palette = "Set1") +
  scale_fill_brewer(name  ="prediction",
                    breaks=c("prediction interval", "credible interval"),
                    palette = "Set1") + 
  geom_point() +
  labs(x = "average ME-2 plasma concentration (ng/mL)",
       y = "average factor Xa inhibition (%)") +
  theme(text = element_text(size = 12),
        axis.text = element_text(size = 12),
        ##          legend.position = c(0.8, 0.25),
        strip.text = element_text(size = 8)) +
  ylim(-10,100)

plot_PPC

```

```{r theEnd}

sessionInfo()

```
